---
title: "Class 09: Candy Mini Project"
author: "Cyrus Shabahang (PID: A19145663)"
format: pdf
toc: TRUE
---

## Background

In today's mini-project we will analyze candy data with ggplot, basic statistics, correlation analysis and principal component analysis methods we have been learning thus far. 

## Data Import 

The data comes as a CSV file from 538. 

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset? 

There are `r nrow(candy)` rows in this dataset. 

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Haribo Twin Snakes", ]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”? (Using dplyr package)

```{r}
library(dplyr)

candy |>
  filter(row.names(candy)=="Kit Kat") |>
  select(winpercent)
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library("skimr")
skim(candy)
```
Winpercent is on a 0-100 scale and is different than the others that are mostly on 0 and 1 indicators. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
table(candy$chocolate)
```

Zero and one are most likely binary indicators. Zero means that the candy does not contain chocolate and one indicates that it does contain chocolate. 

## Exploratory Analysis 

>Q8. Plot a histogram of winpercent values using both base R an ggplot2.

```{r}
hist(candy$winpercent, breaks=20)
```
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=20, fill="lightblue", col="darkgray")
```


>Q9. Is the distribution of winpercent values symmetrical?

No, it is not symmtrical. 


>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
median(candy$winpercent)
```

The center of distribution is slightly below 50% taking the median values, but a mean value of around 50% suggests a right-skewed plot. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Steps to solve this:
1. Find all chocolate candy in the dataset
2. Extract or find their winpercent values
3. Calculate the mean of these values
4. Find all fruity candy
5. Find their winpercent values
6. Calculate their mean vlaue

```{r}
mean(candy$winpercent[candy$chocolate==1])
mean(candy$winpercent[candy$fruity==1])
median(candy$winpercent[candy$chocolate == 1])
median(candy$winpercent[candy$fruity == 1])

```

On average, chocolate candy is ranked higher than fruity candies. This is evident from the mean and median for winpercent both being around 60% for chocolate cnady. 

>Q12. Is this difference statistically significant?

```{r}
choc <- candy$winpercent[as.logical(candy$chocolate)]
fruit <- candy$winpercent[as.logical(candy$fruity)]

t.test(choc, fruit)
```
This difference is statistically significant as indicated by the extremely small p value. A p value below 0.05 tells us that there is statistical signifiance. Chocolate has both a higher mean and median than fruity candies. Additionally, we have a 95% confidence interval. 

## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent), ], n=5)
```
The five least liked candy types in this data set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-candy$winpercent), ], n=5)
```

The top 5 favorite candy types out of this set are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat and Snickers. 

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```
>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r fig.height=10, fig.width=6}
ggplot(candy) +
  aes(winpercent, 
      reorder(rownames(candy), winpercent)) +
  geom_col()
  ylab("")

ggsave("barplot1.png", height=10, width=6)
```
```{r fig.height=10, fig.width=6}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols)

```
>Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets. 

>Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst. 

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
candy$value <- candy$winpercent / candy$pricepercent
candy[order(candy$value, decreasing = TRUE),
      c("pricepercent","winpercent","value")][1:10, ]
```
Tootsie roll midgets are ranked highest in terms of winpercent for the least money. 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
order <- order(candy$pricepercent, decreasing = TRUE)
top5exp <- candy[order, ][1:5, c("pricepercent","winpercent")]
top5exp
top5exp[which.min(top5exp$winpercent), ]
```

The top 5 most expensive candy types are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. Of these, Nik L Nip is the least popular. 

## Exploring the correlation structure

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
```{r}
cij <- cor(candy)
cij["chocolate", "fruity"]
```

```{r}
cij <- cor(candy)
cij[lower.tri(cij)] <- NA
which(cij ==min(cij, na.rm = TRUE), arr.ind = TRUE)
```

Chocolate and fruity are most anti-corrleated. 

>Q23. Similarly, what two variables are most positively correlated?

```{r}
cij <- cor(candy)
max(cij[lower.tri(cij)])
which(cij == max(cij[lower.tri(cij)]), arr.ind = TRUE)

```
The two variables that are most positively correlated are winpercent and chocolate. 

## Principal Component Analysis

```{r}
pca <- prcomp(subset(candy, select = -value), scale. = TRUE)
summary(pca)

```

```{r}
plot(pca$x[,1:2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
library(plotly)
ggplotly(p)
```


