---
title: "Class 8: Breast Cancer Mini Project"
author: "Cyrus Shabahang (PID: A19145663)"
format: pdf
toc: TRUE
---

## Background 

In today's class we will be employing all the R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy data. 

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
```

```{r}
head(wisc.df, 3)
```

> Q1. How many observations are in this dataset?


```{r}
nrow(wisc.df)
```
569 observations

> Q2. How many of the observations have a malignant diagnosis? 

```{r}
sum(wisc.df$diagnosis == "M")
```
212 have a malignant diagnosis

> Q3. How many variables/features in the data are suffixed with _mean? 

```{r}
colnames(wisc.df)
length(grep("mean$", colnames(wisc.df), value = TRUE))
```
10 are suffixed with _mean

We need to remove the `diagnosis` column before we do any further analysis of this dataset - we don't want to pass this to PCA etc. We will save it as a seperate vector that we can use later to compare our findings to those of experts.

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

>Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

44.27%


>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs


>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
## Principal Component Analysis (PCA)

7 PCs


```{r}
wisc.pr <- prcomp(wisc.data, scale. = TRUE)
summary(wisc.pr)
diagnosis <- as.factor(wisc.df$diagnosis)
biplot(wisc.pr)
```
>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot is very difficul to read and there are various overlapping points and labels. 

The main function in base R is called `prcomp()`

The next step in the analysis is to perform principal component analysis (PCA) on wisc.data.

```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

##Execute PCA with the `prcomp()`

```{r}
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```

```{r}
attributes(wisc.pr)
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```


```{r}

ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```
>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

PC1 shows the main separation between malignant and benign. PC2 and PC3 added some spread but not as good as PC1. 


```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var) 

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```
>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]

sort(wisc.pr$rotation[,1], decreasing = TRUE)[1:10]
sort(wisc.pr$rotation[,1], decreasing = FALSE)[1:10]
```
The component of the loading vector is -0.26. Yes there are featuers with larger contributions than this one. 

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?



> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?



>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?


## Hierarchical Clustering 

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
```

## Results of Hierarchical Clustering 

```{r}
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)
```
>Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

The height at which the clustering model has 4 clusters is a height of 19. 


## Combining methods 

The idea here is that I can take my new variables (i.e. the scores on the PCs `wisc.pr$x`) that are better descriptors of the data-set than the original featuers (i.e. the 30 columns in `wisc.data`) and use these as a basis for clustering. 

Clustering on PCA results...

```{r}
pc.dist <- dist(wisc.pr$x[ ,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```
>Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

The method I liked better was ward.D2 because I liked the way it separated the clusters more. 


```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(diagnosis)
```

I can now run `table()` with both of my clustering `grps` and the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

Our cluster "1" has 179 "M" diagnosis 
Our cluster "2" has 333 "B" diagnosis 

179 (True-positive)
24 (False-positive)
333 (True-negative)
33 (False-negative)

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity: TN/(TN+FP)
```{r}
333/(333+24)
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```
>Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

The newly created hclust model diagnoses well. One of the clusers was mostly malignant while the other was mostly benign. There were some wrong groupings but overall it was separated well. 

>Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

The cluster model that didn't use PCA seemed to be more mixed. The PCA with ward.D2 gave a better two grouped split. 


## Prediction

We can use our PCA model for prediction of new un-seen cases: 

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```
>Q16. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize patient 1 for a follow-up since patient 2 looks more benign. 